---
title: "Exploratory Data Analysis of Loan Eligibility Data Set"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Jake Bersabe"
output: 
  html_document:
    css: "style.css"
---

Import data set and libraries:   
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(skimr)
library(ggplot2)
library(printr)

data <- readr::read_csv("train.csv")
glimpse(data)
skim(data)
```



```{r}

# Function to add monthly income column
monthly_payment <- function(loan_amount, loan_amount_term){
  r <- 0.005
  monthly_payment <- (loan_amount/(((1+r)^loan_amount_term)-1))*(((1+r)^loan_amount_term)*r)
  return(monthly_payment)
} 

# Pre-processing pipeline 

data_preprocessed <- data %>%
  mutate(LoanAmount = LoanAmount*1000,
         CombinedIncome = ApplicantIncome + CoapplicantIncome) %>% 
  #select(-ApplicantIncome, -CoapplicantIncome) %>% 
  mutate(MonthlyPayment = monthly_payment(LoanAmount, Loan_Amount_Term),
         PaymentIncomeRatio = MonthlyPayment/CombinedIncome)
  
glimpse(data_preprocessed)
```

### 1. Is there a difference between the median income of applicants with approved loan status and applicants with rejected loan status?   
```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianIncome = median(ApplicantIncome))
```

```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianIncome = median(ApplicantIncome)) %>% 
  ggplot(aes(x = Loan_Status, y = MedianIncome))+
  geom_col(fill = c("#F8766D", "#00BFC4"), alpha = 0.9)+
  geom_text(aes(label=MedianIncome), vjust = -0.5)+
  theme_minimal()
```

>There is not much difference in the median income of both groups.  

### 2. Do applicants with less payment-to-income ratio more likely to be eligible for loan?  
```{r warning=FALSE}
library(ggplot2)
data_preprocessed %>% 
  ggplot(aes(PaymentIncomeRatio, fill = Loan_Status))+
  geom_density(alpha  = 0.7)
```


```{r}
data_preprocessed %>% 
  filter(PaymentIncomeRatio != "") %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MeanPayIncomeRatio = mean(PaymentIncomeRatio, na.rm=T))
```

>Eligible applicants have lower mean of payment-to-income ratio; however, this is not accurate since there are outliers in the data. Median would be the more appropriate measurement to use.  

```{r}
data_preprocessed %>% 
  filter(PaymentIncomeRatio != "") %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianPayIncomeRatio = median(PaymentIncomeRatio, na.rm=T))
```

>There is not much difference in medians between the two groups.   


### 3. Is there a difference in the income of applicants who graduated and those who did not?  
```{r}
library(ggplot2)

# Filtering values greater than 15000 to remove zoom of the outliers and zoom in on the distribution
data_preprocessed %>% 
  filter(ApplicantIncome < 15000) %>% 
  ggplot(aes(y = ApplicantIncome, x = Education, fill = Education))+
  geom_boxplot(alpha = 0.5)

```

```{r}
data_preprocessed %>% 
  group_by(Education) %>% 
  summarise(MedianApplicantIncome = median(ApplicantIncome))
```


>Graduate applicants have higher median income.  

### 4. Do self-employed applicants have lower income?  
```{r}
data_preprocessed %>% 
  filter(Self_Employed != "" & ApplicantIncome < 30000) %>% 
  ggplot(aes(ApplicantIncome, fill = Self_Employed))+
  geom_density(alpha  = 0.6)
```

```{r}
data_preprocessed %>% 
  filter(Self_Employed != "") %>% 
  group_by(Self_Employed) %>% 
  summarise(Count = n(), MedianApplicantIncome = median(ApplicantIncome))
```

>Self-employed applicants actually have higher median income.  

### 5. Are applicants with low monthly payment are more likely to be eligible for loan?   
```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianMonthlyPayment = median(MonthlyPayment, na.rm = T))
```

```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianMonthlyPayment = median(MonthlyPayment, na.rm = T) %>% round(1)) %>% 
  ggplot(aes(Loan_Status, MedianMonthlyPayment))+
  geom_col(fill = c("#F8766D", "#00BFC4"), alpha = 0.85)+
  geom_text(aes(label=MedianMonthlyPayment), vjust = -0.5)+
  theme_minimal()
```

>Eligible applicants have lower median monthly payment.  

### 6. Applicants with higher number of dependents have higher income.   
```{r warning=FALSE}
data_preprocessed %>% 
  ggplot(aes(y = ApplicantIncome, x = Dependents, fill = factor(Dependents)))+
  geom_boxplot(alpha = 0.5)+
  scale_fill_brewer(palette="Set1")
```

```{r}
data_preprocessed %>% 
  filter(Dependents != "") %>% 
  group_by(Dependents) %>% 
  summarise(Count = n(), MedianApplicantIncome = median(ApplicantIncome))
```

>Applicants with 3 or more dependents have the highest median applicant income.  


### 7. How many applicants who are married have dependents?   
```{r}
data_preprocessed %>% 
  filter(Married != "" & Dependents != "") %>% 
  mutate(Dependents = if_else(Dependents > 0, 1,  0)) %>% 
  count(Married, Dependents)
```

```{r}
data_preprocessed %>% 
  filter(Married != "" & Dependents != "") %>% 
  mutate(Dependents = if_else(Dependents > 0, 1,  0)) %>% 
  count(Married, Dependents) %>% 
  ggplot(aes(x = Married, y = n, fill = factor(Dependents)))+
  geom_col(alpha = 0.9)+
  theme_minimal()
```

>Majority of married applicants have dependents; however, married applicants who don't have dependents still hold a big percentage. Majority of unmarried applicants don't have dependents.  

```{r}
data_preprocessed %>% 
  filter(Married == "No" & Dependents != "") %>% 
  mutate(Dependents = if_else(Dependents > 0, 1,  0)) %>% 
  count(Married, Dependents) %>% 
  mutate(Percentage = round(n*100/sum(n),1))
```

### 8. Loan Amount and Applicant Income  

Cluster the applicant income using kmeans clustering algorithm. The clusters are then labeled: "High Income", "Mid Income", and "Low Income".   
```{r}
ApplicantIncomeCluster <- kmeans(data_preprocessed$ApplicantIncome, 3)
data_preprocessed$ApplicantIncomeCluster <- ApplicantIncomeCluster$cluster

data_preprocessed <- data_preprocessed %>% 
  mutate(ApplicantIncomeCluster = if_else(ApplicantIncomeCluster == 1, "High Income",
                                          if_else(ApplicantIncomeCluster == 2, "Mid Income", "Low Income"))) %>% 
  mutate(ApplicantIncomeCluster = ordered(ApplicantIncomeCluster, levels = c("High Income", "Mid Income", "Low Income")))

```



```{r}
data_preprocessed %>% 
  ggplot(aes(y=ApplicantIncome, x=ApplicantIncomeCluster))+
  geom_boxplot(aes(fill = ApplicantIncomeCluster), alpha = 0.4)+
  scale_fill_brewer(palette="Set1")
```

The graph below shows that the kmeans algorithm cleanly clustered the applicants income. 

```{r}
data_preprocessed %>% 
  ggplot(aes(x=Loan_ID, y=ApplicantIncome, color = ApplicantIncomeCluster))+
  geom_point(size = 3, alpha = 0.9)+
  theme_void()+
  scale_color_brewer(palette="Set1")
```


```{r warning=F}
data_preprocessed %>% 
  ggplot(aes(x = ApplicantIncome, y = LoanAmount, color=ApplicantIncomeCluster))+
  geom_point(alpha = 0.8, size = 3)+
  scale_color_brewer(palette="Set1")
```

>Clusters 2 and 3 of applicant income have greater range of loan amount than cluster 1 applicants. 




### 9. Applicants with higher income pay the loan in shorter term
```{r warning=FALSE}
data_preprocessed %>% 
  filter(!is.na(Loan_Amount_Term)) %>% 
  mutate(Loan_Amount_Term = forcats::fct_lump_n(factor(Loan_Amount_Term), 4)) %>% 
  #mutate(Loan_Amount_Term  = factor(Loan_Amount_Term)) %>% 
  ggplot(aes(Loan_Amount_Term, fill = ApplicantIncomeCluster))+
  geom_bar(stat = "count", position = "stack", alpha = 0.9)+
  scale_fill_brewer(palette="Set1")+
  theme_minimal()

```

```{r}
data_preprocessed %>% 
  filter(!is.na(Loan_Amount_Term)) %>% 
  mutate(Loan_Amount_Term = forcats::fct_lump_n(factor(Loan_Amount_Term), 4)) %>% 
  group_by(Loan_Amount_Term, ApplicantIncomeCluster) %>% 
  count(sort = T) %>% 
  arrange(Loan_Amount_Term)
```


>360-month is the favorite loan amoun term across all income clusters.  

### 10. Are most of the applicants with zero co-applicant income unmarried?  
```{r}
data_preprocessed %>% 
  filter(Married != "") %>% 
  mutate(WithCoapplicantIncome = if_else(CoapplicantIncome > 0, 1, 0)) %>% 
  group_by(Married, WithCoapplicantIncome) %>% 
  count() %>% 
  ggplot(aes(x = Married, y = n, fill = factor(WithCoapplicantIncome)))+
  geom_col(position = "dodge", alpha = 0.8)
```

>Most married applicants have co-applicant income and most unmarried applicants don't have co-applicant income.  

### 11. Do male applicants have higher income than female applicants?  
```{r}
data_preprocessed %>% 
  filter(Gender != "") %>% 
  group_by(Gender) %>% 
  summarise(Count = n(), median(ApplicantIncome))
```

```{r}
data_preprocessed %>% 
  filter(Gender != "") %>% 
  ggplot(aes(y=ApplicantIncome, x=Gender, fill = Gender))+
  geom_boxplot(alpha = 0.5)
```

>Apparently, males have slightly higher income than females and male income distribution has wider range.    


