---
title: "Exploratory Data Analysis of Loan Eligibility Data Set"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Jake Bersabe"
output: 
  html_document:
    css: "style.css"
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = F, warning = F, fig.align = 'center')
```


### Import data set and libraries

```{r message=FALSE, warning=FALSE}
library(dplyr) 
library(skimr) 
library(ggplot2)
library(printr) # for tidy tabulations

data <- readr::read_csv("train.csv")
glimpse(data)
skim(data)
```

We have a total of 614 records and 13 columns. We also have some missing values and the numeric variables are not distributed normally.

Let's add some columns which banks use to evaluate the eligibility of loan applicants.

```{r}

# Function to add monthly income column
monthly_payment <- function(loan_amount, loan_amount_term){
  r <- 0.005
  monthly_payment <- (loan_amount/(((1+r)^loan_amount_term)-1))*(((1+r)^loan_amount_term)*r)
  return(monthly_payment)
} 

# Pre-processing pipeline 

data_preprocessed <- data %>%
  mutate(LoanAmount = LoanAmount*1000,
         CombinedIncome = ApplicantIncome + CoapplicantIncome) %>% 
  mutate(MonthlyPayment = round(monthly_payment(LoanAmount, Loan_Amount_Term),2),
         PaymentIncomeRatio = round(MonthlyPayment/CombinedIncome,2))
  
head(data_preprocessed)
```

### 1. Income and Loan Status

Is there a difference between the median income of applicants with approved loan status and applicants with rejected loan status?

```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianIncome = median(ApplicantIncome))
```

```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianIncome = median(ApplicantIncome)) %>% 
  ggplot(aes(x = Loan_Status, y = MedianIncome))+
  geom_col(fill = c("#F8766D", "#00BFC4"), alpha = 0.9)+
  geom_text(aes(label=MedianIncome), vjust = -0.5)+
  theme_minimal()+
  labs(title =  "Loan Status and Income",
       x = "Loan Status",
       y = "Meadian Income")
```

> There is not much difference in the median income of both groups.

### 2. Payment-to-Income Ratio and Loan Status

Do applicants with less payment-to-income ratio more likely to be eligible for loan?

```{r warning=FALSE}
library(ggplot2)
data_preprocessed %>% 
  ggplot(aes(PaymentIncomeRatio, fill = Loan_Status))+
  geom_density(alpha  = 0.7)+
  labs(title =  "Distribution of Payment-Income Ratio According to Loan Status",
       fill = "Loan Status",
       x = "Payment-Income Ratio")+
  theme_minimal()
```

```{r}
data_preprocessed %>% 
  filter(PaymentIncomeRatio != "") %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MeanPayIncomeRatio = mean(PaymentIncomeRatio, na.rm=T))
```

> Eligible applicants have lower mean of payment-to-income ratio; however, this is not accurate since there are outliers in the data. Median would be the more appropriate measurement to use.

```{r}
data_preprocessed %>% 
  filter(PaymentIncomeRatio != "") %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianPayIncomeRatio = median(PaymentIncomeRatio, na.rm=T))
```

> There is not much difference in medians between the two groups.

### 3. Income and Education

Is there a difference in the income of applicants who graduated and those who did not?

```{r}
library(ggplot2)

# Filtering values greater than 15000 to remove zoom of the outliers and zoom in on the distribution
data_preprocessed %>% 
  filter(ApplicantIncome < 15000) %>% 
  ggplot(aes(y = ApplicantIncome, x = Education, fill = Education))+
  geom_boxplot(alpha = 0.7)+
  theme_minimal()+
  labs(title =  "Distribution of Applicant Income According to Education",
       y = "Applicant Income")

```

```{r}
data_preprocessed %>% 
  group_by(Education) %>% 
  summarise(MedianApplicantIncome = median(ApplicantIncome))
```

> Graduate applicants have higher median income.

### 4. Self-employment and Income

Do self-employed applicants have lower income?

```{r}
data_preprocessed %>% 
  filter(Self_Employed != "" & ApplicantIncome < 30000) %>% 
  ggplot(aes(ApplicantIncome, fill = Self_Employed))+
  geom_density(alpha  = 0.6)+
  labs(title =  "Distribution of Applicant Income",
       fill = "Self-employed",
       x = "Applicant Income")+
  theme_minimal()
```

```{r}
data_preprocessed %>% 
  filter(Self_Employed != "") %>% 
  group_by(Self_Employed) %>% 
  summarise(Count = n(), MedianApplicantIncome = median(ApplicantIncome))
```

> Self-employed applicants actually have higher median income.

### 5. Monthly Payment and Loan Status

Are applicants with low monthly payment are more likely to be eligible for loan?

```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianMonthlyPayment = median(MonthlyPayment, na.rm = T))
```

```{r}
data_preprocessed %>% 
  group_by(Loan_Status) %>% 
  summarise(Count = n(), MedianMonthlyPayment = median(MonthlyPayment, na.rm = T) %>% round(1)) %>% 
  ggplot(aes(Loan_Status, MedianMonthlyPayment))+
  geom_col(fill = c("#F8766D", "#00BFC4"), alpha = 0.85)+
  geom_text(aes(label=MedianMonthlyPayment), vjust = -0.5)+
  theme_minimal()+
  labs(title =  "Loan Status and Monthly Payment",
       x = "Loan Status",
       y = "Median Monthly Payment")
```

> Eligible applicants have lower median monthly payment.

### 6. Dependents and Income

Applicants with higher number of dependents have higher income.

```{r warning=FALSE}
data_preprocessed %>% 
  ggplot(aes(y = ApplicantIncome, x = Dependents, fill = factor(Dependents)))+
  geom_boxplot(alpha = 0.7)+
  scale_fill_brewer(palette="Set1")+
  theme_minimal()+
  labs(title =  "Distribution of Applicant Income According to Number of Dependents",
       fill = "Dependents",
       y = "Applicant Income")
```

```{r}
data_preprocessed %>% 
  filter(Dependents != "") %>% 
  group_by(Dependents) %>% 
  summarise(Count = n(), MedianApplicantIncome = median(ApplicantIncome))
```

> Applicants with 3 or more dependents have the highest median applicant income.

### 7. Civil Status and Dependents

How many applicants who are married have dependents?

```{r}
data_preprocessed %>% 
  filter(Married != "" & Dependents != "") %>% 
  mutate(Dependents = if_else(Dependents > 0, 1,  0)) %>% 
  count(Married, Dependents)
```

```{r}
data_preprocessed %>% 
  filter(Married != "" & Dependents != "") %>% 
  mutate(Dependents = if_else(Dependents > 0, 1,  0)) %>% 
  count(Married, Dependents) %>% 
  ggplot(aes(x = Married, y = n, fill = factor(Dependents)))+
  geom_col(alpha = 0.9)+
  theme_minimal()+
  labs(title =  "Civil Status and Dependents",
       y = "Count",
       fill = "Dependents")
```

> Majority of married applicants have dependents; however, married applicants who don't have dependents still hold a big percentage. Majority of unmarried applicants don't have dependents.

```{r}
data_preprocessed %>% 
  filter(Married == "No" & Dependents != "") %>% 
  mutate(Dependents = if_else(Dependents > 0, 1,  0)) %>% 
  count(Married, Dependents) %>% 
  mutate(Percentage = round(n*100/sum(n),1))
```

### 8. Loan Amount and Applicant Income

Cluster the applicant income using kmeans clustering algorithm. The clusters are then labeled: "High Income", "Mid Income", and "Low Income".

```{r}
set.seed(123)
ApplicantIncomeCluster <- kmeans(data_preprocessed$ApplicantIncome, 3)
data_preprocessed$ApplicantIncomeCluster <- ApplicantIncomeCluster$cluster

data_preprocessed <- data_preprocessed %>% 
  mutate(ApplicantIncomeCluster = case_when(ApplicantIncomeCluster  == 3 ~ "High Income",
                                            ApplicantIncomeCluster  == 2 ~ "Mid Income",
                                            ApplicantIncomeCluster  == 1 ~ "Low Income")) %>% 
  mutate(ApplicantIncomeCluster = ordered(ApplicantIncomeCluster, levels = c("High Income", "Mid Income", "Low Income")))

```

```{r}
data_preprocessed %>% 
  ggplot(aes(y=ApplicantIncome, x=ApplicantIncomeCluster))+
  geom_boxplot(aes(fill = ApplicantIncomeCluster), alpha = 0.7)+
  scale_fill_brewer(palette="Set1")+
  theme_minimal()+
  labs(title =  "Distribution of Income According to Applicant Income Cluster",
       y = "Applicant Income",
       x = "Applicant Income Cluster",
       fill = "Cluster")
```

The graph below shows that the kmeans algorithm cleanly clustered the applicants income.

```{r}
data_preprocessed %>% 
  ggplot(aes(x=Loan_ID, y=ApplicantIncome, color = ApplicantIncomeCluster))+
  geom_point(size = 3, alpha = 0.7)+
  theme_void()+
  scale_color_brewer(palette="Set1")
```

```{r warning=F}
data_preprocessed %>% 
  ggplot(aes(x = ApplicantIncome, y = LoanAmount, color=ApplicantIncomeCluster))+
  geom_point(alpha = 0.8, size = 3)+
  scale_color_brewer(palette="Set1")+
  theme_minimal()+
  labs(title = "Applicant Income vs Loan Amount",
       color = "Cluster",
       x = "Applicant Income",
       y = "Loan Amount")
  
```

> Clusters 2 and 3 of applicant income have greater range of loan amount than cluster 1 applicants.

### 9. Income and Loan Term

Applicants with higher income pay the loan in shorter term.

```{r warning=FALSE}
data_preprocessed %>% 
  filter(!is.na(Loan_Amount_Term)) %>% 
  mutate(Loan_Amount_Term = forcats::fct_lump_n(factor(Loan_Amount_Term), 4)) %>% 
  #mutate(Loan_Amount_Term  = factor(Loan_Amount_Term)) %>% 
  ggplot(aes(Loan_Amount_Term, fill = ApplicantIncomeCluster))+
  geom_bar(stat = "count", position = "stack", alpha = 0.9)+
  scale_fill_brewer(palette="Set1")+
  theme_minimal()+
  labs(title = "Loan Amount Term and Applicant Income Cluster",
       fill = "Cluster",
       y = "Count",
       x = "Loan Amount Term")

```

```{r}
data_preprocessed %>% 
  filter(!is.na(Loan_Amount_Term)) %>% 
  mutate(Loan_Amount_Term = forcats::fct_lump_n(factor(Loan_Amount_Term), 4)) %>% 
  group_by(Loan_Amount_Term, ApplicantIncomeCluster) %>% 
  count(sort = T) %>% 
  arrange(Loan_Amount_Term)
```

> 360-month is the favorite loan amoun term across all income clusters.

### 10. Co-applicant Income and Civil Status

Are most of the applicants with zero co-applicant income unmarried?

```{r}
data_preprocessed %>% 
  filter(Married != "") %>% 
  mutate(WithCoapplicantIncome = if_else(CoapplicantIncome > 0, 1, 0)) %>% 
  group_by(Married, WithCoapplicantIncome) %>% 
  count() %>% 
  ggplot(aes(x = Married, y = n, fill = factor(WithCoapplicantIncome)))+
  geom_col(position = "dodge", alpha = 0.8)+
  theme_minimal()+
  labs(title = "Civil Status and Co-applicant",
       subtitle = "This is an attempt to see the relationship between civil status \nand having/not having co-applicant income.")
```

> Most married applicants have co-applicant income and most unmarried applicants don't have co-applicant income.

### 11. Gender and Income

Do male applicants have higher income than female applicants?

```{r}
data_preprocessed %>% 
  filter(Gender != "") %>% 
  group_by(Gender) %>% 
  summarise(Count = n(), median(ApplicantIncome))
```

```{r}
data_preprocessed %>% 
  filter(Gender != "") %>% 
  ggplot(aes(y=ApplicantIncome, x=Gender, fill = Gender))+
  geom_boxplot(alpha = 0.5)+
  theme_minimal()+
  labs(title = "Distribution of Applicant Income According to Gender",
       y = "Applicant Income")
```

> Apparently, males have slightly higher income than females and male income distribution has wider range.
